% 02_nv_sensing.tex — Section 2: NV Sensing Layer
% Phase 2 content migration — Task #193, Cycle 59 NVB-1, 2026-03-20
% Primary source: simulations/nv_sensitivity/nv_sensitivity_sim.tex
% Law 16: quantitative claims carry source locator comments.
% Status: POPULATED (Phase 2, Cycle 59)

\section{NV Sensing Layer}
\label{sec:nv-sensing}

The NV sensing layer converts the magnetic field change from the NbN Meissner transition
into an optically detected frequency shift. This section covers the NV physics background,
the shot-noise-limited sensitivity model, key experimental parameters, and projected
performance across measurement protocols at $T = 12$\,K.

\subsection{NV Physical Background}
\label{sec:nv-physics}

The nitrogen-vacancy (NV) centre in diamond is a spin-1 defect whose $m_s = \pm 1$
ground-state sub-levels are split from $m_s = 0$ by the zero-field splitting
$D/2\pi \approx 2.87$\,GHz at room temperature~\cite{Doherty2013}. % Doherty2013 p.24 ("D_gs reduces to ~2.87 GHz at room temperature")
At $T = 12$\,K, Jacob's system measures $D = 2877.5$\,MHz (This work; CryoLab).

An external magnetic field $B$ shifts the $m_s = 0 \leftrightarrow \pm 1$ transition
frequencies by
\begin{equation}
  \delta f = \gamma_\mathrm{NV} B,
  \qquad \gamma_\mathrm{NV} = \frac{g_e \mu_B}{h} \approx 28.024\;\text{GHz/T}
  \label{eq:freq_shift}
\end{equation}
where $g_e = 2.003$ is the NV electron g-factor and $\mu_B$ is the Bohr
magneton~\cite{Doherty2013}. % Doherty2013 Table 2 (g_gs = 2.0028; Table 1, gamma_NV = 28.024 GHz/T)

Optically detected magnetic resonance (ODMR) measures this frequency shift via the
spin-state-dependent fluorescence contrast~\cite{Barry2020}.
% Barry2020 Eq. (6) p.~10 (ODMR frequency shift; Zeeman splitting formula)

\subsection{Shot-Noise-Limited Sensitivity Formulas}
\label{sec:sensitivity_formulas}

All sensitivity expressions follow Barry et al.\ (2020)~\cite{Barry2020} and represent
theoretical lower bounds. Technical noise (laser intensity noise, MW power fluctuations,
temperature drift of $D$) will in general degrade performance beyond these limits.

\subsubsection{CW-ODMR Sensitivity}
\label{sec:cw_sensitivity}

For continuous-wave ODMR on a Lorentzian resonance, the shot-noise-limited sensitivity
is~\cite{Barry2020}: % Barry2020 Eq. (13) p.~17 (sensitivity eta, general CW-ODMR form)
\begin{equation}
  \eta_\mathrm{CW} = \frac{4}{3\sqrt{3}}
                     \frac{\Delta\nu_\mathrm{FWHM}}
                          {\gamma_\mathrm{NV}\, C_\mathrm{CW}\, \sqrt{R}}
  \label{eq:cw_sensitivity}
\end{equation}
where $\Delta\nu_\mathrm{FWHM}$ is the ODMR linewidth (FWHM), $C_\mathrm{CW}$ is the
fractional ODMR contrast, $R$ is the detected photon rate in s$^{-1}$, and
$\gamma_\mathrm{NV} = 28.024$\,GHz/T~\cite{Doherty2013}. % Doherty2013 Table 2 (gamma_NV)
Jacob's system: $\Delta\nu = 10$\,MHz, $C = 6.5\%$ at 12\,K (This work).

The prefactor $4/(3\sqrt{3})$ arises because the maximum slope of a Lorentzian lineshape
occurs at a detuning $\Delta = \Delta\nu/(2\sqrt{3})$ from line centre, giving a slope
$|dL/df|_\mathrm{max} = 3\sqrt{3}/(4\Delta\nu)$~\cite{Barry2020}. % Barry2020 Sec. III.A (CW-ODMR slope derivation)

\subsubsection{Ramsey Magnetometry Sensitivity}
\label{sec:ramsey_sensitivity}

Ramsey magnetometry uses a free-precession sequence to probe the dephasing time $T_2^*$.
The shot-noise-limited sensitivity is~\cite{Barry2020}: % Barry2020 Eq. (16) p.~18 (Ramsey sensitivity, pulsed free-precession)
\begin{equation}
  \eta_\mathrm{Ramsey} \approx
    \frac{\hbar}{\Delta m_s\, g_e\, \mu_B}
    \cdot \frac{e^{(\tau/T_2^*)^p}}{C_\mathrm{Ramsey}}
    \cdot \sqrt{\frac{t_I + \tau + t_R}{\tau}}
    \cdot \frac{1}{\sqrt{n_\mathrm{avg}}}
  \label{eq:ramsey_sensitivity}
\end{equation}
where $\tau$ is the free-precession time, $t_I$ is the spin initialization time, $t_R$
is the readout time, $p \approx 1$ for inhomogeneous broadening, and
$n_\mathrm{avg} = R\, t_R$ is the average detected photons per cycle. The optimal
sensitivity is achieved at $\tau_\mathrm{opt} = T_2^*$~\cite{Barry2020}.

\textbf{Caveat:} $T_2^*$ has not yet been directly measured in Jacob's system (open
question OQ-SP-01). The simulation estimates $T_2^* \approx 1/(\pi\Delta\nu)
\approx 32$\,ns from the observed linewidth~\cite{Barry2020}. % Barry2020 Sec. III.B (T2* estimation from linewidth)

\subsubsection{Pulsed ODMR Sensitivity}
\label{sec:pulsed_sensitivity}

Pulsed ODMR uses a $\pi/2 - \tau - \pi/2$ sequence. The sensitivity is~\cite{Barry2020}: % Barry2020 Sec. III.C (pulsed ODMR sensitivity)
\begin{equation}
  \eta_\mathrm{pulsed} \approx \frac{8}{3\sqrt{3}}
    \cdot \frac{\hbar}{g_e\, \mu_B}
    \cdot \frac{1}{C_\mathrm{pulsed}\, \sqrt{n_\mathrm{avg}}}
    \cdot \sqrt{\frac{t_I + T_2^* + t_R}{T_2^*}}
  \label{eq:pulsed_sensitivity}
\end{equation}
The prefactor $8/(3\sqrt{3})$ reflects the halved linewidth of pulsed ODMR relative to
CW-ODMR~\cite{Barry2020}.

\subsection{Simulation Parameters and Validation}
\label{sec:sim_validation}

The sensitivity simulation suite (\texttt{NVBolometer/simulations/nv\_sensitivity/})
implements the above formulas and validates against Jacob's measured CW-ODMR sensitivity.

\subsubsection{Experimental Baseline}

\begin{table}[ht]
\centering
\caption{Baseline experimental parameters at 12\,K (CryoLab). ``This work'' entries are
  CryoLab measurements. Literature values cite PDF-in-hand sources.}
\label{tab:nv_params}
\small
\begin{tabular}{@{}p{4.5cm}p{2.5cm}p{2cm}p{4cm}@{}}
\toprule
Parameter & Symbol & Value & Source \\
\midrule
CW-ODMR sensitivity & $\eta_\mathrm{CW}$ & 14\,$\mu$T/$\sqrt{\text{Hz}}$ & This work \\
ODMR linewidth (FWHM) & $\Delta\nu$ & 10\,MHz & This work \\
ODMR contrast & $C$ & 6.5\% & This work \\
Zero-field splitting & $D/2\pi$ & 2877.5\,MHz & This work \\
Operating temperature & $T$ & 12\,K & This work \\
$T_1$ (cryo plateau) & $T_1$ & 230\,ms & This work,~\cite{Jarmola2012} \\ % Jarmola2012 Fig.~3 (T1 vs T; plateau at intermediate T for HPHT NV)
NV gyromagnetic ratio & $\gamma_\mathrm{NV}$ & 28.024\,GHz/T & \cite{Doherty2013} \\ % Doherty2013 Table 2 (g_gs = 2.0028; derived gamma_NV)
Emission band & ZPL at 637\,nm & 637--800\,nm & \cite{Doherty2013} \\ % Doherty2013 Fig.~1(b) p.5 (ZPL at 637 nm; phonon sideband to ~800 nm)
\bottomrule
\end{tabular}
\end{table}

\subsubsection{Derived Simulation Parameters}

\begin{table}[ht]
\centering
\caption{Derived simulation parameters. Not directly measured; clearly flagged.}
\label{tab:derived_params}
\small
\begin{tabular}{@{}p{4.5cm}p{2cm}p{4.5cm}@{}}
\toprule
Parameter & Value & Derivation \\
\midrule
$T_2^*$ (estimated) & $\approx 32$\,ns & $1/(\pi\Delta\nu)$~\cite{Barry2020}
  % Barry2020 Sec. III.B (T2* from linewidth estimate)
  \\
$T_2$ (conservative est.) & $\approx 77$\,ms & $T_1/3$~\cite{Jarmola2012}
  % Jarmola2012 Fig.~3 (T1/T2 ratio for HPHT NV ensemble)
  \\
Implied photon rate & back-calculated & Eq.~\eqref{eq:cw_sensitivity} inverted \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{Validation: Round-Trip Check}

The primary simulation validation tests internal algebraic consistency:
\begin{enumerate}
  \item Back-calculate the implied photon detection rate $R_\mathrm{implied}$ by inverting
    Eq.~\eqref{eq:cw_sensitivity}:
    \begin{equation}
      R_\mathrm{implied} =
      \left(
        \frac{4}{3\sqrt{3}}
        \cdot \frac{\Delta\nu}{\gamma_\mathrm{NV}\, C\, \eta_\mathrm{CW}}
      \right)^2
      \label{eq:back_calc}
    \end{equation}
    using Jacob's values: $\eta_\mathrm{CW} = 14\,\mu\text{T}/\sqrt{\text{Hz}}$,
    $\Delta\nu = 10$\,MHz, $C = 6.5\%$.
  \item Forward-calculate $\eta_\mathrm{check}$ from $R_\mathrm{implied}$ using
    Eq.~\eqref{eq:cw_sensitivity}.
  \item Compute round-trip error $\epsilon = |\eta_\mathrm{check} - \eta_\mathrm{CW}|/\eta_\mathrm{CW}$.
    \textbf{Pass criterion:} $\epsilon < 1\%$.
\end{enumerate}

\textbf{Result:} The round-trip check yields $\epsilon = 0.0000\%$, confirming algebraic
consistency. The validation is not a physical test --- it confirms mathematical self-consistency
of the Barry 2020 formula implementation, not that the formula matches Jacob's system. Physical
validation requires measuring the photon rate independently.
% internal: nv_sensitivity_sim.tex Sec.~5 (Validation result)

\subsection{SiPM Detector Parameters}
\label{sec:sipm-params}

The planned D1 readout detector is the Hamamatsu S14420-3050MG MPPC~\cite{S14420datasheet}.
Key simulation parameters:
% S14420datasheet PDF IN HAND: hardware/sipm_fpga/references/S14420_datasheet.pdf

\begin{table}[ht]
\centering
\caption{S14420-3050MG SiPM parameters used in the sensitivity simulation.}
\label{tab:sipm_sim_params}
\small
\begin{tabular}{@{}p{4.5cm}p{2.5cm}p{5cm}@{}}
\toprule
Parameter & Value & Source \\
\midrule
PDE at 700\,nm & 26\% & \cite{S14420datasheet}, Fig.~3 \\ % S14420datasheet p.3 Fig. right panel (50 µm pitch curve at 700 nm)
PDE at 600\,nm & 40\% & \cite{S14420datasheet} \\ % S14420datasheet p.1 Features
Active area & 7.07\,mm$^2$ & \cite{S14420datasheet} \\
Gain $M$ & $3.6 \times 10^6$ & \cite{S14420datasheet} \\
DCR at 25$^\circ$C & $1.6 \times 10^6$\,s$^{-1}$ & \cite{S14420datasheet} \\ % S14420datasheet p.2 Table (Dark count Typ.)
DCR at 12\,K (estimated) & $\approx 10$\,s$^{-1}$ & Extrapolated from~\cite{Liu2022SiPMLN2} \\ % Liu2022SiPMLN2 (6 orders of magnitude DCR reduction at 77 K)
Breakdown voltage $V_\mathrm{br}$ & 42\,V & \cite{S14420datasheet} \\ % S14420datasheet p.2 Table (V_BR typ at 25°C)
\bottomrule
\end{tabular}
\end{table}

\textbf{Note on cryogenic DCR:} The value of $\approx 10$\,s$^{-1}$ at 12\,K is an
extrapolated estimate. Liu et al.\ (2022)~\cite{Liu2022SiPMLN2} measured a DCR reduction
of approximately 6 orders of magnitude from room temperature to 77\,K for a comparable SiPM
device. The value used here is conservative (assumes somewhat less reduction at 12\,K
vs.\ 77\,K). Direct characterization of the S14420 at 12\,K is an open measurement task.

\subsection{Sensitivity Budget}
\label{sec:sensitivity_budget}

Table~\ref{tab:sensitivity_budget} summarizes the sensitivity budget across five system
configurations from the simulation suite. All entries use the same linewidth and contrast as
Jacob's measurements. Ramsey improvement factors use $T_2^* \approx 32$\,ns (estimated)
and $t_R = 1\,\mu$s readout~\cite{Barry2020}. % Barry2020 Eq. (16) (Ramsey sensitivity at optimal tau)

\begin{table}[ht]
\centering
\caption{Sensitivity budget: CW-ODMR and Ramsey across detector configurations.
  All values from Eqs.~\eqref{eq:cw_sensitivity}--\eqref{eq:ramsey_sensitivity}~\cite{Barry2020}.
  SiPM sensitivity entries are simulation outputs (run \texttt{run\_sensitivity\_analysis.py}).}
\label{tab:sensitivity_budget}
\small
\begin{tabular}{@{}p{5.5cm}p{2.5cm}p{2.5cm}p{2cm}@{}}
\toprule
Configuration
  & $\eta_\mathrm{CW}$ [$\mu$T/$\sqrt{\text{Hz}}$]
  & $\eta_\mathrm{Ramsey}$ [$\mu$T/$\sqrt{\text{Hz}}$]
  & Ramsey/CW \\
\midrule
Current confocal (12\,K) & 14.0 & --- & --- \\
SiPM $\eta_\mathrm{opt} = 2\%$ & (sim output) & (sim output) & (sim output) \\
SiPM $\eta_\mathrm{opt} = 5\%$ & (sim output) & (sim output) & (sim output) \\
SiPM $\eta_\mathrm{opt} = 10\%$ & (sim output) & (sim output) & (sim output) \\
SiPM $\eta_\mathrm{opt} = 20\%$ & (sim output) & (sim output) & (sim output) \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Hahn echo projection:} With $T_2 \approx T_1/3 \approx 77$\,ms (conservative
estimate~\cite{Jarmola2012}) % Jarmola2012 Fig.~3 (T1/T2 ratio for HPHT NV ensemble)
and Hahn-echo readout $t_R = 100\,\mu$s, the simulation projects sensitivity in the low
nT/$\sqrt{\text{Hz}}$ range. \textbf{Critical caveat:} This projection depends entirely on
the unmeasured $T_2$. All Hahn-echo numbers are order-of-magnitude estimates only.
Measuring $T_2$ is open question OQ-SP-01.

\subsection{Known Limitations of the Sensitivity Model}
\label{sec:sensitivity_limits}

The following limitations must be understood before using simulation outputs for design
decisions:
% internal: nv_sensitivity_sim.tex Sec.~7 (Known Limitations, items 1--8)

\begin{enumerate}
  \item \textbf{$T_2^*$ not directly measured.} Estimated as $1/(\pi\Delta\nu) \approx 32$\,ns
    from the observed linewidth~\cite{Barry2020}. Direct measurement via Ramsey fringe decay
    required (OQ-SP-01).
  \item \textbf{$T_2$ not measured.} Conservative estimate $T_2 \approx T_1/3$ follows
    ensemble NV scaling~\cite{Jarmola2012}; actual value unknown. All Hahn-echo projections
    are order-of-magnitude only.
  \item \textbf{Cryogenic SiPM DCR is extrapolated.} The $\approx 10$\,s$^{-1}$ estimate
    at 12\,K is extrapolated from~\cite{Liu2022SiPMLN2}. Characterization of S14420 at
    12\,K is required.
  \item \textbf{NV emission rate is back-calculated.} $R_\mathrm{emit}$ is back-calculated
    from Jacob's sensitivity measurement assuming confocal collection efficiency $\approx 10\%$
    and APD quantum efficiency $\approx 65\%$. Both assumptions are uncalibrated.
  \item \textbf{Contrast is CW-only.} The simulation uses CW contrast for all three
    protocols. Pulsed readout typically achieves $2$--$3\times$ higher contrast than
    CW~\cite{Dreau2011Avoiding}, % Dreau2011Avoiding (power-broadening-avoided pulsed contrast improvement)
    so using $C_\mathrm{CW}$ for Ramsey and pulsed ODMR is a conservative lower bound.
  \item \textbf{Power broadening not modelled.} The linewidth $\Delta\nu = 10$\,MHz is
    taken as an experimental input; power broadening would degrade $\eta_\mathrm{CW}$
    linearly (Eq.~\eqref{eq:cw_sensitivity}).
\end{enumerate}

\subsection{Fluorescence Collection and Diamond Material}
\label{sec:nv-material}

The NV emission at 637--800\,nm is collected by the cryostat confocal objective.
For D0, the confocal beam spot ($\sim\!1\,\mu$m$^2$) defines the sensing volume; no NV
patterning is required. Key collection parameters are addressed in
Section~\ref{sec:oq4}.

Jacob's HPHT diamond has an estimated NV density
$n_\mathrm{NV} \sim 10^{17}$\,cm$^{-3}$, consistent with nitrogen-bath-dominated
dephasing at high NV density, typical of HPHT-grade material~\cite{Barry2020}. % Barry2020 Sec. V (optimal concentration ~10^17 cm^{-3} for bulk NV ensemble magnetometry)
HPHT diamonds commonly achieve $T_2^* \sim 1$--$10\,\mu$s~\cite{Barry2020}. % Barry2020 Table I (T2* range for HPHT, nitrogen-dominated dephasing)
For improved sensitivity in D1, lower-nitrogen CVD diamond with
$n_\mathrm{NV} \sim 10^{16}$--$10^{17}$\,cm$^{-3}$ is preferred;
see Section~\ref{sec:pixel-fab} (Q5).

\subsection{Validation Target Statement}
\label{sec:nv_validation_target}

Per NVB simulation law (CLAUDE.md, Law~10), the sensitivity simulation declares:

\textbf{Validation target:} Reproduce Jacob's measured
$\eta_\mathrm{CW} = 14\,\mu\text{T}/\sqrt{\text{Hz}}$ via the round-trip algebraic
consistency check (Sec.~\ref{sec:sim_validation}).

\textbf{Reference formula:} Barry et al.\ (2020), Eq.~16~\cite{Barry2020}. % Barry2020 Eq. 16 p.~18

\textbf{Result:} Round-trip relative error $= 0.0000\%$ (pass; threshold $< 1\%$).

\textbf{Status:} \textsc{Validated} (algebraic consistency only; physical validation against
independently measured photon rate is still pending).
% internal: nv_sensitivity_sim.tex Sec.~8 (Validation Target Statement, result)
