% 04_sipm_characterization.tex — Section 4: SiPM Characterization & Cryo Behavior
% Phase 2 content migration — Task #178, Cycle 54 NVB-1, 2026-03-20
% Sources: hardware/sipm_noise_characterization.tex,
%          hardware/sipm_cryo_survey.tex,
%          hardware/sipm_vbr_temperature_compensation.tex
% Law 16: quantitative claims carry source locator comments.
% Status: POPULATED (Phase 2)

\section{SiPM Characterization \& Cryo Behavior}
\label{sec:sipm-cryo}

The Hamamatsu S14420-3050MG SiPM is the photon detector for the D1 readout chain.
This section summarizes its room-temperature specifications, cryogenic dark-count-rate
(DCR) behavior, and the critical breakdown-voltage temperature dependence that governs
safe operation at 12\,K.

\subsection{S14420-3050MG Room-Temperature Parameters}

All values from the Hamamatsu datasheet~\cite{S14420datasheet} (kapd1061e, 2023, PDF in hand).
% S14420datasheet p.1-3 (specifications table and PDE figure)

\begin{table}[ht]
\centering
\caption{S14420-3050MG key parameters at 25$^\circ$C~\cite{S14420datasheet}.
  \% S14420datasheet p.2 (electrical characteristics table).}
\label{tab:s14420_params}
\small
\begin{tabular}{@{}p{5cm}p{3cm}p{5cm}@{}}
\toprule
Parameter & Value & Note \\
\midrule
Active area & $3 \times 3$\,mm$^2$ & 3600 microcells \\
Microcell pitch & 50\,$\mu$m & 50\,$\mu$m cell variant \\
PDE at 600\,nm & 40\% & Peak PDE~\cite{S14420datasheet} \\
PDE at 700\,nm (NV emission) & $\sim$25\% & From PDE curve~\cite{S14420datasheet} \\
% S14420datasheet p.3 Fig. right panel (50 um pitch PDE vs wavelength at Vover = 2.6 V)
Breakdown voltage $V_\mathrm{BR}$ & $42 \pm 5$\,V at 25$^\circ$C & \cite{S14420datasheet}
  % S14420datasheet p.2 Table (V_BR min-max)
  \\
DCR (typical) & 1600\,kcps & At $V_\mathrm{op}$, 25$^\circ$C~\cite{S14420datasheet}
  % S14420datasheet p.2 Table (Dark count rate S14420-3050MG Typ.)
  \\
Operating temperature (specified) & $-40$--$+50^\circ$C & \textbf{NOT rated below $-40^\circ$C}~\cite{S14420datasheet}
  % S14420datasheet p.1 Table (Operating temperature range)
  \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Cryogenic DCR: Temperature Model}

Using a Shockley--Read--Hall (SRH) thermal generation model calibrated against published
cryogenic data for ON~Semiconductor J-Series SiPMs (literature calibration;
PDF in hand: \texttt{hardware/sipm\_fpga/references/SiPM\_cryo\_submeV\_arXiv2509.11954.pdf}),
the effective SRH activation energy is $E_a \approx 0.107$\,eV~\cite{SiPMcryoSubmeV2025}.
% SiPMcryoSubmeV2025 (SiPM cryo characterization: DCR vs. T model, activation energy)
% internal: hardware/sipm_noise_characterization.tex Sec.~2 (SRH model, E_a ~ 0.107 eV)

At $T = 12$\,K, the thermal SRH contribution is negligibly small.
The dominant DCR mechanism becomes tunneling-floor current, estimated at
$\lesssim 0.035$\,counts\,s$^{-1}$ for the S14420 active area under low-overvoltage conditions.
% internal: hardware/sipm_noise_characterization.tex Abstract (tunneling floor < 0.035 cps at 12 K)
The project conservative estimate of 10\,cps DCR at 12\,K remains within the
literature-modeled bound and is not inconsistent with higher-overvoltage operation.

\textbf{OPEN GAP (G-GAP-03):} No S14420-specific cryogenic characterization exists
in the published literature. A direct cooldown measurement is required to confirm
the 12\,K DCR, PDE, and breakdown voltage for the S14420 specifically.
% internal: hardware/sipm_noise_characterization.tex Sec.~4 (G-GAP-03 open item)

\subsection{Critical Safety Constraint: Breakdown Voltage vs.\ Temperature}

\textbf{This is the most important pre-D1 measurement required.}

The S14420 breakdown voltage shifts with temperature at approximately
$dV_\mathrm{BR}/dT \approx -47$\,mV/$^\circ$C~\cite{Liu2022SiPMLN2}.
% Liu2022SiPMLN2 (SiPM VBR vs T characterization; ~47-54 mV/K typical for Si APD devices)
% internal: hardware/sipm_vbr_temperature_compensation.tex (VBR(T) model and extrapolation)
Extrapolating from room temperature to 12\,K ($\Delta T = 263$\,K):
\begin{equation}
  V_\mathrm{BR}(12\,\text{K}) \approx 42\,\text{V} - 0.047\,\text{V/K} \times 263\,\text{K}
  \approx 28.6\,\text{V}.
  \label{eq:vbr_cryo}
\end{equation}

\begin{table}[ht]
\centering
\caption{Breakdown voltage and safe operating range at key temperatures.
  Values extrapolated from $dV_\mathrm{BR}/dT \approx -47$\,mV/K~\cite{Liu2022SiPMLN2}.
  \textbf{Operating at room-temperature bias at 12\,K is unsafe.}}
\label{tab:vbr_vs_T}
\small
\begin{tabular}{@{}p{2.5cm}p{2.5cm}p{2.5cm}p{5cm}@{}}
\toprule
$T$ & $V_\mathrm{BR}(T)$ & Safe $V_\mathrm{op}$ & Note \\
\midrule
25$^\circ$C & 42\,V & 44--46\,V & Room-temperature spec~\cite{S14420datasheet} \\
$-40^\circ$C (233\,K) & 52\,V (est.) & 54--56\,V & Manufacturer's low-T limit \\
12\,K & 28.6\,V (est.) & 30--31\,V & \textbf{Lab bias must be reprogrammed before cooldown} \\
\bottomrule
\end{tabular}
\end{table}

\begin{quote}
\textbf{Operating at the room-temperature bias setting (44\,V) at 12\,K would produce
$V_\mathrm{over} \approx 15$\,V --- far above the second-breakdown threshold ---
and must be avoided.} Gate~G-SAFE-01 requires SiPM bias to be reduced to
$\leq\!31$\,V before the system is cooled below 200\,K.
\end{quote}
% internal: hardware/hardware_readiness_report.tex Sec.~4 (safety gate G-SAFE-01)

\subsection{Literature Survey: Comparable SiPMs at Cryogenic Temperature}

Published cryogenic SiPM characterization covers several comparable devices~\cite{SiPMcryoArrays2025,SiPMcryoSubmeV2025,Liu2022SiPMLN2}:
% SiPMcryoArrays2025 (SiPM cryo 40--300 K, gain and noise behavior)
% SiPMcryoSubmeV2025 (sub-meV detector SiPM cryo characterization)
% Liu2022SiPMLN2 (SiPM at LN2 and cryo temperatures: VBR, DCR, gain)
\begin{itemize}
  \item At 77\,K (LN2): DCR drops by $3$--$4$ orders of magnitude vs.\ room temperature;
    gain increases due to lower thermal noise in avalanche region.
  \item At $< 50$\,K: tunneling-floor DCR dominates; values in the 0.01--100\,cps range
    depending on device and overvoltage.
  \item The S14420 is not characterized below $-40^\circ$C in any published study.
    \textbf{Direct measurement at 12\,K is required (G-GAP-03).}
\end{itemize}
