% 06_hardware_readiness.tex — Section 6: Hardware Readiness Assessment
% Phase 2 content migration — Task #178, Cycle 54 NVB-1, 2026-03-20
% Source: hardware/hardware_readiness_report.tex
% Law 16: quantitative claims carry source locator comments.
% Status: POPULATED (Phase 2)

\section{Hardware Readiness Assessment}
\label{sec:hardware-readiness}

This section is the formal D0 hardware gate: \textbf{Are we ready to run the Meissner--Zeeman
demonstration?} \textbf{Short answer: not yet, due to two blocking items: (1)~NbN film not yet
deposited; (2)~SiPM cryogenic bias characterization not yet measured.}
% internal: hardware/hardware_readiness_report.tex Abstract (gate status summary)

\subsection{Hardware In Hand and Verified}

\begin{longtable}{@{}p{4cm}p{2.5cm}p{6cm}@{}}
\toprule
\textbf{Component} & \textbf{Status} & \textbf{Verification} \\
\midrule
\endhead
532\,nm confocal laser & IN HAND &
  Actively used; $D = 2877.5$\,MHz confirmed at 12\,K (This work) \\
Montana cryostat, 11\,K base & IN HAND &
  ODMR data at 12--250\,K; 11--70\,K operating range (This work) \\
NV diamond (HPHT ensemble) & IN HAND &
  Contrast 6--7.5\%; $T_1 \approx 230$\,ms; $\eta \approx 14\,\mu$T/$\sqrt{\text{Hz}}$
  at 12\,K (This work) \\
ADF4351 MW synthesizer & IN HAND &
  35\,MHz--4.4\,GHz; $+5$\,dBm; SPI~\cite{ADF4351datasheet}
  % ADF4351datasheet p.4 Table 1 (frequency range and output power)
  \\
NdFeB bias magnet & IN HAND &
  Reaches $\sim$46\,G; used in 10.6--210\,G ODMR campaigns (This work) \\
S14420-3050MG SiPM & IN HAND &
  PDE = 40\% at 600\,nm; $V_\mathrm{BR} = 42$\,V at 25$^\circ$C~\cite{S14420datasheet}
  % S14420datasheet p.2-3
  \\
MAX40658AUT+ TIA & IN HAND & 18.3\,k$\Omega$; 360--520\,MHz BW~\cite{MAX40658datasheet}
  % MAX40658datasheet p.3 Table
  \\
ADCMP601 comparator & IN HAND & 3.5\,ns prop. delay~\cite{ADCMP601datasheet}
  % ADCMP601datasheet p.4 Table 1
  \\
Nexys A7-100T FPGA & IN HAND &
  XC7A100T; replaces discontinued Arty A7-100T~\cite{NexysA7refmanual}
  % NexysA7refmanual p.1 Table
  \\
ADL5375 IQ modulator & IN HAND &
  Datasheet and eval board in hand; D1 upgrade path, not needed for D0~\cite{ADL5375datasheet} \\
ADL5321 RF driver amp & IN HAND &
  Gain 11.7\,dB at 2877\,MHz~\cite{ADL5321datasheet}
  % ADL5321datasheet Table 2 (S21 at 2877 MHz interpolated)
  \\
Mather~(1982) nonequilibrium NEP theory & IN HAND &
  PDF obtained 2026-03-19~\cite{Mather1982}.
  % Mather1982 (ILL request fulfilled; nonequilibrium corrections to bolometer NEP)
  \\
\midrule
\caption{D0 hardware in hand and verified.}
\label{tab:hw_in_hand}
\endlastfoot
\bottomrule
\end{longtable}

\subsection{Hardware Blockers and Gates}

\begin{longtable}{@{}p{3.5cm}p{2cm}p{7cm}@{}}
\toprule
\textbf{Item} & \textbf{Status} & \textbf{Notes} \\
\midrule
\endhead
NbN thin film on sapphire (200\,nm, $T_c \geq 15$\,K) &
  \textbf{NEEDED} &
  \textbf{Primary D0 blocker.} Film must be deposited before D0 can run.
  UT Arlington Shimadzu NRC deposition is the recommended path
  (see Section~\ref{sec:nbn-sourcing}). \\[4pt]
S14420 cryogenic bias characterization (G-GAP-03) &
  \textbf{NEEDED} &
  $V_\mathrm{BR}(T)$ must be measured by direct cooldown before D1 SiPM bias is applied
  at 12\,K. \textbf{Gate G-SAFE-01: bias must be reduced before cooling below 200\,K.}
  % internal: hardware/hardware_readiness_report.tex Sec.~4 (G-SAFE-01 gate)
  \\[4pt]
NbN film $T_c$ verification & \textbf{NEEDED} &
  As-deposited film must have $T_c \geq 15$\,K confirmed by resistance measurement
  before D0 experiment can be interpreted.
  % Kamlapure2010 Table I (T_c ~ 15.87 K for epitaxial NbN; 200 nm target)
  \\
\midrule
\caption{D0 hardware blockers and safety gates.}
\label{tab:hw_blockers}
\endlastfoot
\bottomrule
\end{longtable}

\subsection{Critical Path to D0}

\begin{enumerate}
  \item \textbf{Email \texttt{nanofab@uta.edu}} to confirm NbN sputtering target and N$_2$
    reactive gas availability at UT Arlington Shimadzu NRC.
  \item If confirmed: submit deposition request; estimated turnaround 4--8 weeks (process
    development + deposition run). % internal: hardware/nbn_film_sourcing.md Sec.~5 (lead time estimate)
  \item Deposit and measure $T_c$ by resistance vs.\ temperature.
  \item Once $T_c \geq 15$\,K confirmed: run D0-Easy experiment
    (blanket NbN film + existing diamond + existing cryostat + existing ODMR setup).
  \item \textbf{Separately} (D1): characterize S14420 $V_\mathrm{BR}(T)$ by direct cooldown;
    program cryogenic bias before any SiPM power-up below 200\,K.
\end{enumerate}

\subsection{NEP Model Status}

The bolometer NEP model~\cite{Richards1994} is derived from first principles:
% Richards1994 Eq. (18) p.8 (NEP_phonon = sqrt(4 gamma k_B T^2 G))
$\mathrm{NEP}_\mathrm{phonon}^2 = 4\gamma k_B T^2 G_{ep}$ at $\gamma = 1$, with
$G_{ep} = 5\Sigma V T^4 \approx 4.7 \times 10^{-5}$\,W/K at 12\,K for the D0 pixel.
% internal: simulations/bolometer_thermal_model.tex (G_ep = 4.7e-5 W/K at 12 K)
Nonequilibrium corrections~\cite{Mather1982} (up to 30\% phonon noise reduction,
60\% Johnson noise reduction) are quantified in the companion simulation
(\texttt{bolometer\_thermal\_model.tex}).
