% 09_oq5_energy_resolution.tex — Section 9: OQ-5 Energy Resolution (Threshold Detector)
% Phase 2 full-body migration — Task #210, Cycle NVB, 2026-03-21
% Source: open_questions/oq5_energy_resolution.tex (rewrite 2026-03-20; authoritative)
% Law 16: quantitative claims carry source locator comments.
% Status: POPULATED (Phase 2 full body) — full body extracted from oq5_energy_resolution.tex
%
% NOTE: \bibliography{} block removed — master document handles bibliography.
%       All \cite{}, \needref{} retained. siunitx macros (\si{}, \SI{}) retained.
%       The master preamble loads amsmath, amssymb but NOT siunitx — see compile notes.
%
% ARCHIVE: Previous section content (summary) archived as 09_oq5_energy_resolution_phase2_task210_archive_2026-03-21.tex
%
% COMPILE NOTE: This section uses \si{} and \SI{} from the siunitx package.
%   Add \usepackage{siunitx} to nvbolometer_master.tex preamble if not already present.

\section{OQ-5: Threshold Energy of the NbN Bolometer Pixel}
\label{sec:oq5}

\subsection{Detector Paradigm: Threshold vs.\ Calorimeter}
\label{sec:oq5_paradigm}

The distinction between a \emph{threshold detector} and a \emph{linear
calorimeter} is fundamental to interpreting energy sensitivity
\cite{Richards1994}. % Richards1994 Sec. II.H, p. 17 (superconducting bolometers operate at the SC transition)

\subsubsection{Linear Calorimeter (Previous Model --- Incorrect for D0)}

In the linear calorimeter paradigm, the bolometer operates at a fixed
temperature on the resistive transition slope. A small energy deposition
$\delta E$ produces a proportional temperature perturbation
$\delta T = \delta E / C$, which is read out via a change in resistance.
The minimum detectable energy is set by the noise floor:
\begin{equation}
  E_\mathrm{res} = \mathrm{NEP}_\mathrm{total} \times \sqrt{\tau}
  \quad [\text{J}].
  \label{eq:oq5_E_res_calorimeter}
\end{equation}
This is the correct model for a transition-edge sensor (TES) operated with
electrothermal feedback that holds the film at a stable bias point on the
transition~\cite{Richards1994}. % Richards1994 Sec. II.H, p. 17 (TES operational description)
The Wollman/Hattori optical TES, for example, achieves $67$\,meV
FWHM energy resolution at $T_c = 115$\,mK using this
approach~\cite{Wollman2022OpticalTES}. % Wollman2022OpticalTES Abstract p. 1, Table 1 p. 4

\textbf{The previous version of this document applied
Eq.~(\ref{eq:oq5_E_res_calorimeter}) to D0 and obtained
$E_\mathrm{res}^\mathrm{ph} = 48$\,eV. This was an error.} D0 does
not operate as a TES. It has no electrothermal feedback loop, no SQUID
readout, and no voltage bias to maintain a stable operating point. D0 uses a
CW-ODMR readout of an adjacent NV centre, which detects the magnetic field
change when the NbN film transitions from the superconducting (Meissner) to
the normal state.

\subsubsection{Threshold Detector (Correct Model for D0)}

In a threshold detector, the superconducting film sits at
$T_\mathrm{op} < T_c$ in the fully Meissner-screened state. When an energy
deposition heats the pixel above $T_c$, the film transitions to the normal
state and the Meissner screening collapses. The NV centre detects this
collapse via the change in local magnetic field (OQ-1 mechanism).

The minimum energy required to trigger the transition is the \emph{thermal
threshold energy}: the heat needed to raise the pixel from $T_\mathrm{op}$
to $T_c$,
\begin{equation}
  \boxed{
  E_\mathrm{th} = \int_{T_\mathrm{op}}^{T_c} C_\mathrm{total}(T)\,dT
  \quad [\text{J}].
  }
  \label{eq:oq5_E_th_def}
\end{equation}
This is \emph{not} a noise-limited quantity; it is a thermodynamic property
of the pixel. An energy deposition $E_\mathrm{dep} < E_\mathrm{th}$ cannot
trigger the transition regardless of readout sensitivity.

This is analogous to the threshold model used in TES-based particle
detectors, where the relevant quantity is the energy needed to drive the
sensor above the superconducting
transition~\cite{APPEC2021DirectDetection}. % APPEC2021DirectDetection Sec. 4.6.1, p. 28 (cryogenic bolometer energy thresholds 30--70 eV in sub-K systems)

\subsection{Thermal Model: Heat Capacity of the D0 Pixel}
\label{sec:oq5_thermal_model}

The D0 pixel consists of two thermally coupled layers:
\begin{enumerate}
  \item \textbf{NbN thin film} (superconducting screen): $200$\,nm
    thick, footprint $1.5 \times 5$\,$\mu$m$^2$.
  \item \textbf{Ge absorber}: $1$\,$\mu$m thick, same footprint.
\end{enumerate}

\noindent Both layers share the same footprint area
$A = 7.5$\,$\mu$m$^2$.

\subsubsection{NbN Electronic Heat Capacity (Sommerfeld Model)}
\label{sec:oq5_C_NbN}

For a metal in the normal state, the electronic heat capacity per unit
volume follows the Sommerfeld (free-electron) model~\cite{Richards1994}: % Richards1994 Appendix p. 23 (C_el = gamma*T for metals)
\begin{equation}
  c_e = \gamma_V\,T \quad [\text{J\,m}^{-3}\,\text{K}^{-1}],
  \label{eq:oq5_sommerfeld}
\end{equation}
where $\gamma_V$ is the volumetric Sommerfeld coefficient
[$\text{J\,m}^{-3}\,\text{K}^{-2}$]. In the Drude model,
$\gamma_V = (\pi^2/3)\,k_B^2\,N(0)$, where $N(0)$ is the electronic
density of states at the Fermi level (both
spins)~\cite{Richards1994}. % Richards1994 Appendix p. 23 (gamma = pi^2 k^2 m N / hbar^2 (3pi^2 N/V)^{2/3})

Sidorova et al.~\cite{Sidorova2020} report the single-spin density of
states $N(0) = 3.0$--$5.0 \times 10^{28}$\,eV$^{-1}$\,m$^{-3}$ for
disordered NbN thin films on Si/SiO$_2$, extracted from Hall and transport
measurements. % Sidorova2020 Table III, p. 11 (N(0) column: 3.02--4.98 e28 eV^-1 m^-3)
Using the Drude relation with total (both-spin) DOS:
\begin{equation}
  \gamma_V = \frac{\pi^2}{3}\,k_B^2 \times 2\,N(0).
  \label{eq:oq5_gamma_drude}
\end{equation}
For $N(0) = 4.0 \times 10^{28}$\,eV$^{-1}$\,m$^{-3}$
($= 2.50 \times 10^{47}$\,J$^{-1}$\,m$^{-3}$), this gives
$\gamma_V \approx 313$\,J\,m$^{-3}$\,K$^{-2}$.
% Sidorova2020 Table III p.11 (N(0) = 4.0e28 eV^-1 m^-3, mid-range of reported values)

\needref{NbN thin-film Sommerfeld coefficient $\gamma_V$ from direct
calorimetric measurement. Literature suggests $\gamma_V \approx
200$--$600$\,J\,m$^{-3}$\,K$^{-2}$ depending on film disorder and
stoichiometry.}

\textbf{Note on the superconducting state:} For $T < T_c$, the BCS
electronic heat capacity is exponentially suppressed by the gap:
$C_e^\mathrm{BCS} \sim C_e^n \exp(-\Delta/k_B T)$. At $T/T_c = 0.76$
(i.e.\ $T_\mathrm{op} = 12$\,K), the suppression factor is
$\exp(-1.76\,T_c/T) \approx 0.1$. However, for computing
$E_\mathrm{th}$ we integrate through the transition: the film
\emph{must} pass through $T_c$ to trigger detection. Near $T_c$ the
electronic heat capacity rises sharply to its normal-state value (and
overshoots due to the BCS jump). Using the normal-state Sommerfeld
model throughout the integral is therefore a \emph{conservative
(lower-bound)} approximation, since the BCS heat capacity jump at $T_c$
adds energy that must be
supplied~\cite{Richards1994}. % Richards1994 Appendix p. 23 (BCS heat capacity rises sharply at T_c)

The total NbN film heat capacity is
\begin{equation}
  C_\mathrm{NbN}(T) = \gamma_V\,V_\mathrm{NbN}\,T,
  \label{eq:oq5_C_NbN}
\end{equation}
with $V_\mathrm{NbN} = L \times W \times d = 5.0\,\mu\text{m} \times
1.5\,\mu\text{m} \times 200\,\text{nm} =
1.50 \times 10^{-18}$\,m$^3$. % internal: NVB_TRAINING.md Sec. 4 (Demonstrator pixel geometry)

\subsubsection{Ge Absorber Heat Capacity (Debye $T^3$ Law)}
\label{sec:oq5_C_Ge}

The Ge absorber is an insulator (intrinsic Ge at cryogenic $T$), so its
heat capacity is purely phononic. In the Debye low-temperature limit
($T \ll \Theta_D$)~\cite{Richards1994}: % Richards1994 Appendix p. 22 (C = 234 N k_B (T/Theta)^3 per atom)
\begin{equation}
  C_\mathrm{Ge}(T)
  = \frac{12\,\pi^4}{5}\,n_\mathrm{Ge}\,V_\mathrm{Ge}\,k_B
    \left(\frac{T}{\Theta_D^\mathrm{Ge}}\right)^3
  \quad [\text{J/K}],
  \label{eq:oq5_C_Ge}
\end{equation}
where $n_\mathrm{Ge}$ is the atomic number density, $V_\mathrm{Ge}$ is the
absorber volume, and $\Theta_D^\mathrm{Ge} = 374$\,K is the Debye
temperature of Ge. % Richards1994 Appendix p. 22 (Ge(363 K) listed; standard value 374 K from Flubacher1959)

The Ge absorber parameters are:
\begin{align}
  d_\mathrm{Ge}   &= 1.0\,\mu\text{m}, \notag \\
  V_\mathrm{Ge}   &= 1.5\,\mu\text{m} \times 5.0\,\mu\text{m}
                      \times 1.0\,\mu\text{m}
                    = 7.50 \times 10^{-18}\,\text{m}^3, \notag \\
  \rho_\mathrm{Ge} &= 5323\,\text{kg\,m}^{-3}, \quad
  M_\mathrm{Ge}    = 72.63\,\text{g\,mol}^{-1}, \notag \\
  n_\mathrm{Ge}   &= \frac{\rho_\mathrm{Ge}\,N_A}{M_\mathrm{Ge}}
                    = 4.41 \times 10^{28}\,\text{m}^{-3}.
  \label{eq:oq5_Ge_params}
\end{align}
% internal: standard Ge material properties (rho, M from periodic table/CRC handbook)

\subsubsection{Total Pixel Heat Capacity}

The total heat capacity of the NbN+Ge pixel is
\begin{equation}
  C_\mathrm{total}(T) = C_\mathrm{NbN}(T) + C_\mathrm{Ge}(T)
  = \gamma_V V_\mathrm{NbN}\,T
    + \frac{12\pi^4}{5}\,n_\mathrm{Ge}\,V_\mathrm{Ge}\,k_B
      \left(\frac{T}{\Theta_D}\right)^3.
  \label{eq:oq5_C_total}
\end{equation}

Table~\ref{tab:oq5_heat_cap} compares the two contributions at selected
temperatures. The Ge absorber dominates the total heat capacity at all
relevant temperatures by a factor of $3$--$6\times$, because its volume is
$5\times$ larger than the NbN film and the Debye $T^3$ phonon heat capacity
grows rapidly with temperature.

\begin{table}[ht]
\centering
\caption{Heat capacities of the D0 pixel components at selected
temperatures. $C_\mathrm{NbN}$ uses the Sommerfeld model with
$\gamma_V = 600$\,J\,m$^{-3}$\,K$^{-2}$ (conservative upper estimate).
$C_\mathrm{Ge}$ uses the Debye $T^3$ law with $\Theta_D = 374$\,K.
Source: Eqs.~(\ref{eq:oq5_C_NbN}), (\ref{eq:oq5_C_Ge}).}
\label{tab:oq5_heat_cap}
\small
\begin{tabular}{@{}llllll@{}}
\toprule
$T$ [K] & $C_\mathrm{NbN}$ [J/K] & $C_\mathrm{Ge}$ [J/K] &
$C_\mathrm{total}$ [J/K] & $C_\mathrm{Ge}/C_\mathrm{NbN}$ \\
\midrule
12.0 & $1.08 \times 10^{-14}$ & $3.53 \times 10^{-14}$ &
  $4.61 \times 10^{-14}$ & $3.3$ \\
  % internal: Python calculation verified by hand
15.0 & $1.35 \times 10^{-14}$ & $6.89 \times 10^{-14}$ &
  $8.24 \times 10^{-14}$ & $5.1$ \\
15.5 & $1.40 \times 10^{-14}$ & $7.61 \times 10^{-14}$ &
  $9.00 \times 10^{-14}$ & $5.5$ \\
15.7 & $1.41 \times 10^{-14}$ & $7.90 \times 10^{-14}$ &
  $9.32 \times 10^{-14}$ & $5.6$ \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Key observation:} The $1\,\mu$m Ge absorber contributes
$\sim\!80\%$ of the total pixel heat capacity. If the goal is to minimise
$E_\mathrm{th}$, reducing the Ge thickness is the single most effective
design lever.

\subsection{Results: Threshold Energy vs.\ Operating Temperature}
\label{sec:oq5_results}

\subsubsection{Analytical Integration}

The integrals in Eq.~(\ref{eq:oq5_E_th_def}) evaluate analytically for both
components:
\begin{align}
  E_\mathrm{th}^\mathrm{NbN}
    &= \int_{T_\mathrm{op}}^{T_c} \gamma_V V_\mathrm{NbN}\,T\,dT
     = \frac{\gamma_V V_\mathrm{NbN}}{2}
       \left(T_c^2 - T_\mathrm{op}^2\right),
  \label{eq:oq5_E_th_NbN} \\[4pt]
  E_\mathrm{th}^\mathrm{Ge}
    &= \int_{T_\mathrm{op}}^{T_c}
       \frac{12\pi^4}{5}\,n_\mathrm{Ge}\,V_\mathrm{Ge}\,k_B
       \left(\frac{T}{\Theta_D}\right)^3 dT
     = \frac{12\pi^4\,n_\mathrm{Ge}\,V_\mathrm{Ge}\,k_B}
            {20\,\Theta_D^3}
       \left(T_c^4 - T_\mathrm{op}^4\right),
  \label{eq:oq5_E_th_Ge} \\[4pt]
  E_\mathrm{th}^\mathrm{total}
    &= E_\mathrm{th}^\mathrm{NbN} + E_\mathrm{th}^\mathrm{Ge}.
  \label{eq:oq5_E_th_total}
\end{align}

\subsubsection{Threshold Energy Table}

Using $T_c = 15.7$\,K (project
standard~\cite{Kamlapure2010}), % Kamlapure2010 (NbN T_c ~ 15.87 K for epitaxial films); internal: NVB_TRAINING.md Sec. 8 (T_c = 15.7 K design value)
$\gamma_V = 600$\,J\,m$^{-3}$\,K$^{-2}$
(\needref{NbN $\gamma_V$ direct measurement}; estimate, see
Sec.~\ref{sec:oq5_limitations}),
and the Ge parameters from Eq.~(\ref{eq:oq5_Ge_params}), we compute
$E_\mathrm{th}$ at four operating temperatures.

\begin{table}[ht]
\centering
\caption{Threshold energy $E_\mathrm{th}$ for the D0 pixel
($T_c = 15.7$\,K) as a function of operating temperature
$T_\mathrm{op}$. ``NbN only'' = no Ge absorber.
``NbN+Ge'' = full pixel with $1\,\mu$m Ge absorber.
All energies computed from Eqs.~(\ref{eq:oq5_E_th_NbN})--(\ref{eq:oq5_E_th_total}).}
\label{tab:oq5_E_th}
\small
\begin{tabular}{@{}lllll@{}}
\toprule
$T_\mathrm{op}$ [K] &
$E_\mathrm{th}^\mathrm{NbN}$ [eV] &
$E_\mathrm{th}^\mathrm{Ge}$ [eV] &
$E_\mathrm{th}^\mathrm{total}$ [eV] &
Status \\
\midrule
12.0  & $2.88 \times 10^5$ & $1.28 \times 10^6$ & $1.56 \times 10^6$ &
  D0 operating point \\
  % internal: Python calculation — E_NbN = 287893 eV, E_Ge = 1275559 eV
15.0  & $6.04 \times 10^4$ & $3.23 \times 10^5$ & $3.83 \times 10^5$ &
  $T_c - T_\mathrm{op} = 0.7$\,K \\
  % internal: Python calculation — E_NbN = 60365 eV, E_Ge = 322937 eV
15.5  & $1.75 \times 10^4$ & $9.68 \times 10^4$ & $1.14 \times 10^5$ &
  $0.2$\,K below $T_c$ \\
  % internal: Python calculation — E_NbN = 17528 eV, E_Ge = 96803 eV
15.68 & $1.76 \times 10^3$ & $9.85 \times 10^3$ & $1.16 \times 10^4$ &
  $0.02$\,K below $T_c$ \\
  % internal: Python calculation — E_NbN = 1763 eV, E_Ge = 9848 eV
$\geq 15.7$ & --- & --- & --- &
  Not detectable \\
\bottomrule
\end{tabular}
\end{table}

\noindent \textbf{Key results from Table~\ref{tab:oq5_E_th}:}

\begin{itemize}
  \item At $T_\mathrm{op} = 12$\,K (the nominal D0 operating point),
    the threshold energy is $E_\mathrm{th} \approx 1.56$\,MeV
    (NbN+Ge) or $\approx 288$\,keV (NbN only).
    This is an \emph{MeV-scale} threshold --- D0 can detect alpha
    particles ($\sim 5$\,MeV) but not keV-scale X-rays.
    % internal: Python calculation verified
  \item The Ge absorber contributes $\sim 80\%$ of $E_\mathrm{th}$ at all
    operating temperatures. Eliminating or thinning the Ge absorber is the
    most direct route to lowering the detection threshold.
  \item Operating at $T_\mathrm{op} = 15.5$\,K ($0.2$\,K below
    $T_c$) reduces $E_\mathrm{th}$ by $\sim 14\times$ relative to
    $12$\,K, reaching $\sim 114$\,keV (NbN+Ge).
  \item At $T_\mathrm{op} \geq T_c = 15.7$\,K, the film is already
    in the normal state; no threshold crossing is possible.
\end{itemize}

\subsubsection{Comparison: Old ``48\,eV'' Result vs.\ New Threshold Model}

The previous version of this document reported
$E_\mathrm{res}^\mathrm{ph} = 48$\,eV as the ``phonon-noise-limited
energy resolution.'' This quantity was derived from
\begin{equation}
  E_\mathrm{res}^\mathrm{ph}
  = \sqrt{4\,k_B\,T^2\,C}
  \quad [\text{J}],
  \label{eq:oq5_old_NEP}
\end{equation}
which is the thermal fluctuation limit for a linear
calorimeter~\cite{Richards1994}. % Richards1994 Eq. (17), p. 8 (thermal fluctuation <(Delta u)^2> = k_B T^2 C)
Equation~(\ref{eq:oq5_old_NEP}) answers the question: ``What is the smallest
energy perturbation detectable \emph{on top of} thermal fluctuations at
temperature $T$?'' This is the correct question for a TES or microcalorimeter
with continuous readout.

For D0, the correct question is: ``How much energy must be deposited to push
the film \emph{across} $T_c$?'' The answer ---
$E_\mathrm{th} = \int C\,dT$ --- is $10^3$--$10^7\times$ larger than the
old $48$\,eV result, depending on $T_\mathrm{op}$.

\textbf{The ``$48$\,eV'' figure is hereby retired and must not appear
in any D0 sensitivity discussion.} The correct D0 sensitivity metric is
$E_\mathrm{th}$ from Table~\ref{tab:oq5_E_th}.

\subsection{Readout Speed Mismatch: Thermal Recovery vs.\ CW-ODMR Integration}
\label{sec:oq5_readout_mismatch}

Even if an energy deposition exceeds $E_\mathrm{th}$ and triggers the
superconducting-to-normal transition, D0 must \emph{detect} that the
transition occurred. The detection mechanism is CW-ODMR of the adjacent NV
centre, which senses the change in local magnetic field when Meissner
screening collapses. There is a fundamental speed mismatch between the
thermal dynamics and the NV readout:

\begin{enumerate}
  \item \textbf{Thermal recovery timescale.}
    After the NbN film is heated above $T_c$, it cools back below $T_c$
    with a time constant $\tau = C/G_{ep}$. At $T = 12$\,K, the
    electron-phonon thermal conductance
    $G_{ep} = 5\,\Sigma\,V_\mathrm{NbN}\,T^4 \sim
    4.7 \times 10^{-5}$\,W/K (using
    $\Sigma \approx 3 \times 10^8$\,W\,m$^{-3}$\,K$^{-5}$, see
    \cite{Richards1994} Sec.~II.J),
    % Richards1994 Sec. II.J, p. 19 (hot-electron regime G_ep = 5*Sigma*V*T^4)
    and the quasiparticle recombination time in NbN is
    $\tau_{ep} \sim 10$--$18$\,ps at
    $10$\,K~\cite{Sidorova2020}. % Sidorova2020 Table III, p. 11 (tau_e-ph(10K) = 11.9--17.5 ps)
    The NbN film can return to the superconducting state within
    $\sim\!\mathrm{ns}$ timescales.

  \item \textbf{CW-ODMR integration timescale.}
    A CW-ODMR measurement of the NV centre ODMR spectrum requires sweeping
    the microwave frequency across the resonance while collecting
    photoluminescence. Typical integration times are
    $\sim 0.1$--$10$\,s per spectrum, depending on required
    signal-to-noise ratio, NV brightness, and collection
    efficiency.
    \needref{CW-ODMR integration time estimate for single NV at cryogenic
    temperature.}

  \item \textbf{The mismatch.}
    The thermal recovery is $\sim 10^{9}$--$10^{10}\times$ faster than the
    ODMR integration. A single threshold-crossing event (particle impact
    deposits $E > E_\mathrm{th}$, film goes normal, then recovers to SC
    within nanoseconds) would appear as a transient glitch in the NV readout
    that is far shorter than the integration window.
\end{enumerate}

\subsubsection{Does the Threshold Crossing Cause a Persistent Change?}

For D0 to function as a threshold detector with CW-ODMR readout, one of the
following must hold:

\begin{enumerate}
  \item[(a)] \textbf{Persistent normal state.} The energy deposition heats
    the film above $T_c$ and the film \emph{remains} in the normal state
    (latched). This requires either: (i) Joule heating from bias current
    sustains the normal state (thermal runaway), or (ii) the cooling power
    is insufficient to bring the film back below $T_c$ before the next
    energy deposition.

  \item[(b)] \textbf{Pulsed readout.} The ODMR measurement is replaced by a
    pulsed NV protocol that can detect transient magnetic field changes on
    nanosecond timescales (e.g.\ Ramsey interferometry, spin echo). This
    would require substantial redesign of the readout scheme.

  \item[(c)] \textbf{Statistical detection.} Repeated energy depositions
    (e.g.\ from a steady particle flux) cause the film to spend a fraction
    of time in the normal state, shifting the time-averaged ODMR signal.
    The CW-ODMR measurement then detects the \emph{average} normal-state
    occupancy.

  \item[(d)] \textbf{Calorimetric mode after all.} If the film does not
    latch in the normal state and the readout cannot resolve individual
    events, D0 may need to operate as a calorimeter (measuring temperature
    rise via the NV ZFS shift, not via threshold crossing). In this case,
    the NEP-based energy resolution from the original model becomes
    relevant again --- but with the caveat that the readout noise is set by
    the NV ODMR sensitivity, not by the bolometer's intrinsic thermal noise.
\end{enumerate}

\textbf{This readout mismatch is flagged as an open question.} The D0
operating mode (threshold vs.\ calorimeter) depends critically on whether a
persistent state change occurs after the threshold crossing. This must be
resolved before D0 sensitivity can be definitively stated. The Meissner--Zeeman
demonstration (OQ-1) will provide the first experimental data on the
timescale of the NV response to a superconducting transition.

\subsection{Comparison to Relevant Energy Scales}
\label{sec:oq5_comparison}

\begin{table}[ht]
\centering
\caption{D0 detection capabilities at $T_\mathrm{op} = 12$\,K
(NbN+Ge pixel, $E_\mathrm{th} \approx 1.56$\,MeV) and at
$T_\mathrm{op} = 15.5$\,K ($E_\mathrm{th} \approx 114$\,keV).
NbN-only pixel values are $\sim 5\times$ lower.}
\label{tab:oq5_comparison}
\small
\begin{tabular}{@{}lllp{5cm}@{}}
\toprule
Signal & $E_\mathrm{dep}$ & Detectable? & Notes \\
\midrule
Alpha particle (5 MeV) & $5$\,MeV & \textbf{YES} (12\,K) &
  $E_\mathrm{dep} \gg E_\mathrm{th}$ \\
Fission fragment & $\sim 80$\,MeV & \textbf{YES} &
  Massive surplus \\
keV X-ray (1\,keV) & $1$\,keV & \textbf{NO} (12\,K); marginal (15.5\,K, NbN-only) &
  NbN-only $E_\mathrm{th}(15.5) = 17.5$\,keV \\
IR photon ($1$\,eV) & $1.24$\,eV & \textbf{NO} &
  $10^6\times$ below threshold \\
CEvNS recoil ($\sim$keV) & $\lesssim 1$\,keV & \textbf{NO} &
  Requires sub-K
  operation~\cite{Akimov2017COHERENT} \\
  % Akimov2017COHERENT Sec. II p. 3 Eq. (1)
DM candidate ($\sim$eV) & $\sim 1$\,eV & \textbf{NO} &
  Requires sub-K + sub-nm film~\cite{APPEC2021DirectDetection} \\
  % APPEC2021DirectDetection Sec. 4.6.1 p. 28
\bottomrule
\end{tabular}
\end{table}

\noindent D0 at $12$\,K is a \emph{MeV-scale} threshold detector,
sensitive to alpha particles and other heavily ionising radiation but
insensitive to X-rays, optical photons, or sub-keV particle physics signals.
This is consistent with the primary D0 science goal (Meissner--Zeeman
demonstration, OQ-1), which does not require energy-resolved particle
detection.

\subsection{Model Limitations and Uncertainty Budget}
\label{sec:oq5_limitations}

\begin{enumerate}

\item \textbf{Sommerfeld coefficient $\gamma_V$ is uncertain by
  $\sim 2\times$.}
  The task specification uses $\gamma_V \approx 600$\,J\,m$^{-3}$\,K$^{-2}$.
  Our Drude estimate from the Sidorova~\cite{Sidorova2020} density of states % Sidorova2020 Table III p. 11
  gives $\gamma_V \approx 313$\,J\,m$^{-3}$\,K$^{-2}$ (using
  $N(0) = 4.0 \times 10^{28}$\,eV$^{-1}$\,m$^{-3}$). The discrepancy
  is within the expected range for disordered NbN: stoichiometry,
  granularity, and electron mass renormalisation all affect $\gamma_V$.
  We use $\gamma_V = 600$\,J\,m$^{-3}$\,K$^{-2}$ as a conservative
  (higher) estimate; the NbN contribution to $E_\mathrm{th}$ could be up
  to $\sim 2\times$ lower. Since Ge dominates the total heat capacity, this
  uncertainty affects $E_\mathrm{th}^\mathrm{total}$ by only
  $\sim 20$--$30\%$.
  \needref{NbN thin-film $\gamma_V$ from calorimetric measurement on
  200\,nm film.}

\item \textbf{Ge Debye temperature uses bulk value.}
  We use $\Theta_D^\mathrm{Ge} = 374$\,K (bulk crystalline Ge).
  Thin-film or amorphous Ge may have a reduced Debye temperature,
  which would \emph{increase} $C_\mathrm{Ge}$ and therefore
  \emph{increase} $E_\mathrm{th}$.
  % Richards1994 Appendix p. 22 (Ge Debye temp listed as 363 K; standard value 374 K from Flubacher1959)

\item \textbf{BCS heat capacity jump neglected.}
  At $T_c$, the BCS theory predicts a heat capacity discontinuity
  $\Delta C / C_n = 1.43$ (weak-coupling BCS). For NbN (strong-coupling),
  the jump ratio can be larger ($\sim 2$). This means the actual energy
  required to push the film \emph{through} $T_c$ is somewhat larger than
  our integral with the normal-state Sommerfeld model predicts.
  Our $E_\mathrm{th}$ values are therefore \emph{lower bounds}.
  % Richards1994 Appendix p. 23 (BCS heat capacity jump discussion)

\item \textbf{Thermal coupling between NbN and Ge layers.}
  We assume perfect thermal coupling: a phonon deposited in the Ge
  instantly thermalises with the NbN. In reality, the Kapitza boundary
  resistance at the NbN/Ge interface may partially decouple the two
  layers, especially on sub-nanosecond timescales. If the layers are
  thermally decoupled, $E_\mathrm{th}$ for triggering the NbN transition
  is determined by $C_\mathrm{NbN}$ alone (the ``NbN-only'' column in
  Table~\ref{tab:oq5_E_th}), which is $\sim 5\times$
  lower~\cite{Richards1994}. % Richards1994 Appendix p. 23 (boundary resistance B/T^3)

\item \textbf{Nonequilibrium phonon spectrum.}
  A particle impact produces a nonequilibrium phonon distribution, not a
  uniform temperature rise. The fraction of deposited energy that
  thermalises in the NbN film (``phonon collection efficiency'') depends
  on the phonon ballistic length, film-substrate interface
  transmission~\cite{Sidorova2020}, % Sidorova2020 Sec. IV.B p. 17 (acoustic mismatch model, phonon escape time)
  and detector geometry. Our threshold model implicitly assumes 100\%
  energy thermalisation in the pixel.

\item \textbf{Readout sensitivity not modelled.}
  The threshold energy $E_\mathrm{th}$ is a thermodynamic quantity. Whether
  D0 can \emph{detect} a threshold crossing depends on the NV readout
  sensitivity, which is a separate open question (OQ-1 and Sec.~\ref{sec:oq5_readout_mismatch}).

\item \textbf{Old ``$48$\,eV'' value was physically wrong, not just
  numerically inaccurate.}
  The NEP-based $48$\,eV result from the previous version was a
  correct calculation of the thermal fluctuation limit for a hypothetical
  NbN calorimeter. The error was in applying it to D0, which is not a
  calorimeter. The $48$\,eV number itself is mathematically correct
  given the parameters used; it simply answers the wrong question.

\end{enumerate}

\subsection{Summary and Conclusions}
\label{sec:oq5_conclusions}

\begin{table}[ht]
\centering
\caption{Summary: OQ-5 Threshold Energy Analysis for D0 (corrected model).}
\label{tab:oq5_summary}
\small
\begin{tabular}{@{}lll@{}}
\toprule
Item & Result & Confidence \\
\midrule
Detector paradigm & \textbf{Threshold} (not calorimeter) & Firm \\
$E_\mathrm{th}$ at 12\,K (NbN+Ge) & $\sim 1.56$\,MeV &
  $\pm 2\times$ ($\gamma_V$ uncertainty) \\
$E_\mathrm{th}$ at 12\,K (NbN only) & $\sim 288$\,keV &
  $\pm 2\times$ \\
$E_\mathrm{th}$ at 15.5\,K (NbN+Ge) & $\sim 114$\,keV &
  $\pm 2\times$ \\
$E_\mathrm{th}$ at 15.68\,K (NbN+Ge) & $\sim 11.6$\,keV &
  $\pm 2\times$ \\
Ge dominance of heat capacity & $\sim 80\%$ of total & Firm \\
Readout mismatch & \textcolor{red}{\textbf{OPEN}} &
  Sub-ns recovery vs.\ s-scale ODMR \\
Old ``48\,eV'' result & \textbf{RETIRED} & Wrong detector paradigm \\
\bottomrule
\end{tabular}
\end{table}

\noindent \textbf{Answer to OQ-5 (corrected):} The NbN Demonstrator pixel
D0 is a threshold detector, not a linear calorimeter. The minimum energy
required to trigger a superconducting-to-normal transition is
$E_\mathrm{th} = \int_{T_\mathrm{op}}^{T_c} C(T)\,dT$, which evaluates to
$\sim 1.56$\,MeV at $T_\mathrm{op} = 12$\,K for the full
NbN+Ge pixel. The previously reported ``$48$\,eV phonon-noise-limited
energy resolution'' applied the wrong detector paradigm and is retired. D0
at 12\,K is sensitive to alpha particles and fission fragments but not to
keV-scale X-rays, eV-scale photons, or sub-eV particle physics signals.

The critical open question is whether the threshold crossing produces a
persistent (or at least readout-resolvable) state change, given the
$\sim 10^{10}\times$ mismatch between thermal recovery time
($\sim\!\mathrm{ns}$) and CW-ODMR integration time ($\sim\!\mathrm{s}$). If
no persistent change occurs, D0 may need to operate in calorimetric mode
(sensing time-averaged temperature shifts via the NV ZFS), in which case the
sensitivity metric reverts to a readout-noise-limited figure --- but one
determined by NV ODMR sensitivity, not by the NbN thermal noise floor.
