% 10_oq4_epifluorescence.tex — Section 10: OQ-4 Epifluorescence Beamline
% Phase 2 content migration — Task #178, Cycle 54 NVB-1, 2026-03-20
% Source: open_questions/oq4_epifluorescence_beamline.tex (authoritative)
% Law 16: quantitative claims carry source locator comments.
% Status: POPULATED (Phase 2) — design requirements summary.
%
% NOTE: Authoritative source is oq4_epifluorescence_beamline.tex.
%       This section presents design requirements and key options.

\section{OQ-4: Epifluorescence Beamline}
\label{sec:oq4}

This section covers the D1 epifluorescence optical beamline design for cryogenic NV
ensemble readout. Full analysis is in \texttt{oq4\_epifluorescence\_beamline.tex}.

\subsection{Design Requirements}

\begin{table}[ht]
\centering
\caption{D1 epifluorescence beamline requirements summary.
  % internal: oq4\_epifluorescence\_beamline.tex Sec.~2 (requirements table)
  Items marked [PI DECISION REQUIRED] need a PI decision before hardware can be ordered.}
\label{tab:oq4_requirements}
\small
\begin{tabular}{@{}p{3.5cm}p{3cm}p{6cm}@{}}
\toprule
Parameter & Required value & Notes \\
\midrule
Excitation $\lambda$ & 532\,nm & Nd:YAG doubled, NV off-resonant excitation~\cite{Rondin2014}
  % Rondin2014 p.4 Fig.1 (NV excitation via 532 nm phonon sideband)
  \\
Collection band & 637--800\,nm & NV$^-$ ZPL at 637\,nm + phonon sideband~\cite{Doherty2013}
  % Doherty2013 Fig.~1(b) p.5 (ZPL at 637 nm, phonon sideband to ~800 nm)
  \\
Excitation rejection & OD $>$ 5 at 532\,nm & Dichroic + notch filter combination \\
Operating temperature & 4--12\,K & Jacob's Montana cryostat \\
NV target & Ensemble (not single-NV) & Relaxes NA requirement; longer WD acceptable \\
Architecture choice & \needref{PI decision required} & In-cryostat vs.\ external objective (see below) \\
Detector & SiPM or sCMOS & \needref{Jacob decides}; different coupling requirements \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Objective Architecture Decision}

\textbf{The critical choice is whether the objective sits inside or outside the cryostat.}
This is the single most consequential architectural decision.

\subsubsection{Architecture A: Objective Inside Cryostat}

\textbf{Advantages:} No window limits NA or working distance; achieves NA 0.65--0.82
at sample; no aberration from a cryostat window.

\textbf{Disadvantages:} Objective must survive 4\,K thermal cycling (standard microscope
objectives contain lubricants that fail at 4\,K); thermal drift during temperature ramp;
focus adjustment requires piezo scanner inside cryostat.

\textbf{Recommended objective (in-cryostat):} Attocube LT-APO/LWD/VIS/0.64 ---
NA 0.65; 5.0\,mm working distance; cryo-rated.
\needref{AttocubeLTAPOLWD\_spec --- datasheet not yet downloaded.}

\subsubsection{Architecture B: Objective External to Cryostat}

\textbf{Advantages:} No cryogenic shock to objective; standard room-temperature optics
can be used; easier focus adjustment.

\textbf{Disadvantages:} NA limited by cryostat window aperture and distance;
aberration from window (vacuum viewport or cold window).
\needref{Cryostat window aperture and sample-to-window distance from Montana Instruments spec.}

\subsection{Filter Chain}

The dichroic and emission filter must:
\begin{itemize}
  \item Pass: 637--800\,nm (NV$^-$ emission band~\cite{Doherty2013})
    % Doherty2013 Fig.~1(b) p.5 (ZPL at 637 nm, sideband to ~800 nm)
  \item Reject: 532\,nm at OD $>$ 5 (dichroic + notch)
  \item Typical choice: Thorlabs DMLP650 longpass dichroic + Semrock LP03-532RE edge filter
    \needref{ThorlabsDMLP650\_spec; SemrockLP03-532RE\_spec (datasheets not yet downloaded).}
\end{itemize}

\subsection{Collection Efficiency Model}

Collection efficiency for an NV ensemble as a function of objective NA~\cite{Rondin2014}:
% Rondin2014 Sec.~IV (collection efficiency and far-field emission pattern)
\begin{equation}
  \eta_\mathrm{coll} \approx \frac{1}{2}\left(1 - \cos\theta_\mathrm{max}\right),
  \quad \theta_\mathrm{max} = \arcsin(\mathrm{NA}/n_\mathrm{diamond}),
  \label{eq:collection_efficiency}
\end{equation}
where $n_\mathrm{diamond} = 2.4$ is the diamond refractive index.

\begin{table}[ht]
\centering
\caption{Collection efficiency vs.\ objective NA for NV emission in diamond ($n = 2.4$).
  % internal: oq4\_epifluorescence\_beamline.tex Table (collection efficiency table)
  Values from Eq.~(\ref{eq:collection_efficiency}).}
\label{tab:collection_efficiency}
\small
\begin{tabular}{@{}p{4.5cm}p{2.5cm}p{2.5cm}p{3cm}@{}}
\toprule
Objective & NA & $\eta_\mathrm{coll}$ & Note \\
\midrule
Attocube LT-APO/LWD & 0.65 & 11.1\% & In-cryostat; 5\,mm WD \\
Attocube LT-APO & 0.82 & 21.4\% & In-cryostat; 0.55\,mm WD \\
Nikon Plan Apo 40$\times$ & 0.95 & 34.1\% & External; standard air objective \\
\bottomrule
\end{tabular}
\end{table}

\subsection{OQ-4 Status}

\textbf{Status: OPEN.} The in-cryostat vs.\ external objective decision requires
a PI decision (\needref{Jacob's decision: objective architecture}).
The filter chain specifications are established; datasheets must be obtained
before hardware procurement (\needref{Thorlabs DMLP650 and Semrock LP03-532RE datasheets}).
