% 11_nbn_sourcing.tex — Section 11: NbN Film Sourcing & D0 Path
% Phase 2 full-body migration — Task #210, Cycle NVB, 2026-03-21
% Source: hardware/nbn_film_sourcing.md (Task #107, NVB, 2026-03-19; authoritative)
%         Converted from markdown to LaTeX body format per task specification.
% Law 16: quantitative claims carry source locator comments (URL inline per nbn_film_sourcing.md).
% Status: POPULATED (Phase 2 full body) — full markdown-to-LaTeX conversion of nbn_film_sourcing.md
%
% ARCHIVE: Previous section content archived as 11_nbn_sourcing_phase2_task210_archive_2026-03-21.tex
%
% NOTE: No \documentclass, \usepackage, \begin{document} — fragment only.
%       Prices gathered 2026-03-19 --- verify before purchasing.
%       URLs cited inline per Law 16 source locator requirement.
%       No \bibliography block — master handles bibliography.

\section{NbN Film Sourcing \& D0 Path}
\label{sec:nbn-sourcing}

The D0 milestone requires an NbN superconducting thin-film on sapphire substrate with
$T_c \geq 15.7$\,K. This section documents the sourcing analysis and recommended
procurement path. Prices gathered 2026-03-19 --- verify before purchasing.
% internal: hardware/nbn_film_sourcing.md (Task #107, NVB, 2026-03-19)

\subsection{Vendor Table}
\label{sec:nbn_vendors}

\begin{table}[ht]
\centering
\caption{NbN thin-film sourcing options evaluated 2026-03-19. Prices are preliminary;
  verify before purchasing.
  % internal: hardware/nbn\_film\_sourcing.md Sec.~1 (vendor table)
  All URLs accessed 2026-03-19; market prices change.}
\label{tab:nbn_vendors}
\small
\begin{tabular}{@{}p{2.5cm}p{3cm}p{1.8cm}p{6cm}@{}}
\toprule
Vendor & Product & $T_c$ spec & Note \\
\midrule
2D Semiconductors & NbN thin film (PLD) on SiO$_2$/Si (standard); custom sapphire on request &
  $\sim$12\,K &
  \textbf{Below 15\,K target.} Custom sapphire substrate on request.
  \url{https://2dsemiconductors.com/NbN-thin-film/}. \$590/sample (2026-03-19). \\[4pt]
AJA International & Contract reactive sputtering on customer substrate &
  Not listed &
  Same AJA ORION class as UT NRC. Contact: \texttt{topgun@ajaint.com}.
  \url{https://www.ajaint.com/deposition-services}. \\[4pt]
STAR Cryoelectronics & Custom LTS/HTS foundry services; NbN not standard &
  N/A &
  Primary process is Nb/Al-AlOx/Nb trilayers. NbN not listed as standard.
  \url{https://starcryo.com/foundry-services/}. \\[4pt]
SeeQC & 40\,nm NbN on 150\,mm high-resistivity Si (internal SFQ process) &
  Not for sale &
  Internal NbN process on Si; not available for external substrate customers.
  \url{https://seeqc.com/foundry-services}. \\[4pt]
Heeger Materials & NbN sputtering targets and NbN powder; sapphire substrates separately &
  N/A (target) &
  Useful if UT Arlington has sputtering capability.
  \url{https://heegermaterials.com/sputtering-targets/463-niobium-nitride-sputtering-target.html}.
  Target cost $\sim$\$200--500. \\[4pt]
Goodfellow & Niobium nitride materials; thin-film form on request &
  Not published &
  High-purity NbN for SC electronics R\&D; thin film on sapphire may be available as custom order.
  \url{https://www.goodfellow.com/usa/resources/niobium-nitride-nbn-material-information/}. \\
\bottomrule
\end{tabular}
\end{table}

\subsection{UT Arlington Shimadzu Institute NRC --- Feasibility Assessment}
\label{sec:nbn_nrc}

\textbf{Facility:} Shimadzu Institute Nanotechnology Research Center (NRC), UT Arlington,
College of Engineering. \url{https://shimadzuinstitute.org/centers/nanotechnology-research-center/}
Contact: \texttt{nanofab@uta.edu} | Phone: 817-272-1536

\textbf{Deposition equipment confirmed on-site:}
\begin{itemize}
  \item Kurt J.\ Lesker (KJL) Lab 18 Sputtering System
  \item AJA ATC ORION Series UHV Sputtering System
  \item Home-Built Sputter System
\end{itemize}
The facility has at least three magnetron sputtering systems, one of which is an AJA ATC ORION UHV platform
--- the same class of system documented in the peer-reviewed literature for NbN deposition by reactive DC
magnetron sputtering (Nb target + N$_2$/Ar reactive gas). Published process parameters for achieving
$T_c \geq 16$\,K with 200\,nm NbN on sapphire by RF magnetron sputtering at substrate temperatures of
$\sim$150$^\circ$C or above are well-established
(source: \url{https://ieeexplore.ieee.org/document/10697750/} --- substrate temp 150$^\circ$C $\to T_c = 16.58$\,K reported).

\textbf{What is NOT yet confirmed:}
\begin{enumerate}
  \item Whether an Nb sputtering target (or stoichiometric NbN target) is
    already installed or available.
  \item Whether the reactive gas (N$_2$) delivery is configured on any of their systems.
  \item Whether external/industry users from a UT faculty research group can get independent access
    or must collaborate with NRC staff.
\end{enumerate}

\textbf{Practical assessment:}
Jacob's group is at UT Arlington Physics. The NRC is an interdisciplinary facility
``open to scientists within and outside of the University.'' Onboarding requires:
project description email to \texttt{nanofab@uta.edu}, user application form,
safety training, equipment-specific training. No NbN-specific barrier has been identified.

\textbf{Recommended action:} Email \texttt{nanofab@uta.edu} with the following query:
\begin{enumerate}
  \item Is an Nb sputtering target (or NbN target) available for reactive sputtering?
  \item Is N$_2$ reactive gas plumbed into the AJA ORION or KJL system?
  \item What is the user fee rate (per-hour or per-run) for sputtering?
  \item Is there prior NbN experience on-site, or will this be a first-run process development?
\end{enumerate}

Nb sputtering targets are commercially available ($\sim$\$200--500 depending on purity/size;
see Heeger Materials:
\url{https://heegermaterials.com/sputtering-targets/463-niobium-nitride-sputtering-target.html}).

\subsection{Academic Deposition Collaboration Options}
\label{sec:nbn_academic}

Several university nanofabs have published NbN deposition capability and may accept
external collaboration:

\begin{table}[ht]
\centering
\caption{University nanofabs with published NbN capability.
  % internal: hardware/nbn\_film\_sourcing.md Sec.~3.1 (academic collaboration table)
  Reserve for if UT Arlington NRC cannot support NbN.}
\label{tab:nbn_academic}
\small
\begin{tabular}{@{}p{4cm}p{5cm}p{4cm}@{}}
\toprule
Institution & Known NbN capability & Contact route \\
\midrule
Cornell CNF (Center for Nanoscale Systems) &
  AJA sputtering systems; NbN for SNSPD work published from CNF users.
  \url{https://www.cnf.cornell.edu/} &
  External user program available \\
MIT Lincoln Laboratory &
  Established NbN SNSPD fabrication line; high-quality ultrathin NbN on sapphire.
  AlN buffer + NbN on sapphire demonstrated with $T_c > 15$\,K &
  Primarily internal; collaborations via PI agreements \\
University at Buffalo / SUNY SAMRI &
  NbN on 300\,mm Si via Applied Materials PVD; AlN buffer + NbN on sapphire demonstrated.
  \url{https://pmc.ncbi.nlm.nih.gov/articles/PMC10707461/} &
  Contact SUNY SAMRI directly \\
UT Austin Texas Materials Institute &
  Thin film deposition capabilities; geographic proximity to UTA.
  \url{https://tmi.utexas.edu/} &
  No NbN-specific publication found; worth querying \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Practical notes on academic collaborations:}
\begin{itemize}
  \item Academic collaborations typically require an MOU or materials transfer agreement (1--4 weeks).
  \item Many nanofabs charge external academic users $\sim$\$100--300/hr for sputtering.
  \item For a simple blanket film deposition (no patterning), the run itself is short ($\sim$1--2 hours);
    cost is primarily process development if the system hasn't run NbN before.
  \item Cornell CNF has the most accessible external user program of the listed options.
\end{itemize}

\subsection{Draft Purchase Specification}
\label{sec:nbn_spec}

\begin{table}[ht]
\centering
\caption{NbN thin-film purchase specification for D0 milestone.
  % internal: hardware/nbn\_film\_sourcing.md Sec.~4 (draft spec)
  All Law~16 quantitative claims cite source URL inline.}
\label{tab:nbn_spec}
\small
\begin{tabular}{@{}p{4cm}p{9cm}@{}}
\toprule
Parameter & Specification \\
\midrule
Substrate & $c$-plane (0001) sapphire (Al$_2$O$_3$), double-side polished (preferred);
  $c$-plane preferred for thermal conductivity and NV diamond proximity.
  $R$-plane acceptable as secondary option. \\
Film material & Niobium nitride (NbN), $\delta$-phase (cubic, B1 crystal structure) \\
Film thickness & $\sim$200\,nm $\pm$ 20\% \\
  & [Rationale: thick enough for robust bulk-like $T_c$; thin enough for
    proximity to NV diamond surface in future device geometries] \\
$T_c$ & $\geq 15.0$\,K (minimum acceptable); target $\geq 15.7$\,K \\
  & [Bulk NbN $T_c \sim 16$\,K; 200\,nm films on sapphire at optimized
    conditions achieve $T_c = 16.58$\,K
    (source: \url{https://ieeexplore.ieee.org/document/10697750/})] \\
% Source: https://ieeexplore.ieee.org/document/10697750/ (T_c = 16.58 K at 200 nm, 150°C substrate)
Transition width & $\Delta T_c < 1$\,K (10--90\% resistance criterion) \\
  & [Narrow transition = uniform film; required for sharp Meissner onset] \\
Substrate size & $\geq 5 \times 5$\,mm$^2$ \\
Sheet resistance & 10--100\,$\Omega$/sq at room temperature \\
  & [For 200\,nm film; RRR ($R_{300\text{K}}/R_{20\text{K}}$) $> 1$ indicates metallic behavior] \\
Surface quality & $R_a < 1$\,nm RMS (double-side polished sapphire substrate) \\
Optional quality & XRD rocking curve FWHM $< 0.5^\circ$ for (200) NbN peak
  (confirms crystalline quality for epitaxial or polycrystalline film) \\
Delivery format & As-deposited (unpackaged wafer piece in cleanroom packaging);
  no patterning required for D0 demonstration \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Recommended Path}
\label{sec:nbn_path}

\begin{enumerate}
  \item \textbf{Primary: UT Arlington NRC in-house deposition.}
    Jacob is at UT Arlington --- no shipping, no MOU delay, direct access and process control.
    The NRC has AJA ORION UHV sputtering --- the correct tool class for NbN deposition.
    Process literature is clear: reactive DC sputtering of Nb in N$_2$/Ar on sapphire at 150$^\circ$C
    achieves $T_c = 16.58$\,K (source: \url{https://ieeexplore.ieee.org/document/10697750/}).
    Cost: primarily process development time + materials ($\sim$\$500--1,000 estimated total).
    % internal: hardware/nbn_film_sourcing.md Sec.~5 (primary recommendation rationale)
    \textbf{Gating action:} Email \texttt{nanofab@uta.edu} to confirm N$_2$ reactive gas and
    Nb target availability. Estimated turnaround to first film: 4--8 weeks (training + process
    development run).

  \item \textbf{Secondary: AJA International contract deposition.}
    If UT Arlington NRC cannot support NbN:
    AJA International (Hingham, MA) offers contract deposition on customer-supplied substrates.
    Uses same AJA systems as NRC; reactive N$_2$/Ar sputtering of Nb is their documented capability.
    Jacob supplies sapphire substrate; AJA deposits NbN to spec.
    Contact: \texttt{topgun@ajaint.com}; Phone: 781-545-7365. Lead time: estimate 3--6 weeks.

  \item \textbf{Tertiary: 2D Semiconductors (expedient proof-of-concept only).}
    2D Semiconductors sells NbN on SiO$_2$/Si at \$590/sample (in stock).
    $T_c$ confirmed $\sim$12\,K on their standard substrate --- this is \textbf{below our
    $T_c \geq 15$\,K requirement}.
    Suitable only as a proof-of-concept if $T_c \geq 12$\,K is acceptable for initial ODMR testing.
    URL: \url{https://2dsemiconductors.com/NbN-thin-film/}.
    \textbf{Not recommended for D0.}

  \item \textbf{Do not pursue for D0:}
    \begin{itemize}
      \item \textbf{SeeQC:} Internal NbN process on Si, not available for external substrate customers.
      \item \textbf{STAR Cryoelectronics:} NbN not their standard process; primarily Nb/Al-AlOx/Nb.
      \item \textbf{Academic collaborations (Cornell, MIT LL):} Viable but slower;
        reserve for if UTA NRC fails.
    \end{itemize}
\end{enumerate}

\subsection{Key Technical Claims (Law~16 Compliance)}
\label{sec:nbn_claims}

\begin{table}[ht]
\centering
\caption{NbN sourcing key facts with source URLs (Law~16).
  % internal: hardware/nbn\_film\_sourcing.md Sec.~6 (law 16 table)
  All URLs accessed 2026-03-19.}
\label{tab:nbn_claims}
\small
\begin{tabular}{@{}p{5cm}p{2.5cm}p{6cm}@{}}
\toprule
Claim & Value & Source \\
\midrule
NbN 200\,nm on sapphire, RF magnetron sputtering, 150$^\circ$C substrate $\to T_c$ &
  16.58\,K & \url{https://ieeexplore.ieee.org/document/10697750/} \\
NbN films on unoptimized sapphire (bare substrate) &
  $T_c$ 9--10\,K & \url{https://www.mdpi.com/2079-6412/15/5/513} \\
NbN/AlN buffer/sapphire, $\sim$5\,nm NbN &
  $T_c$ up to 13.3\,K & \url{https://arxiv.org/html/2511.13917v1} \\
AJA ORION used for NbN reactive sputtering &
  Published & \url{https://www.sciencedirect.com/science/article/abs/pii/S0925838820322891} \\
UT Arlington NRC equipment & AJA ORION UHV confirmed &
  \url{https://shimadzuinstitute.org/centers/nanotechnology-research-center/} \\
2D Semiconductors NbN price (SiO$_2$/Si) &
  \$590 (2026-03-19) & \url{https://2dsemiconductors.com/NbN-thin-film/} \\
SeeQC NbN: 40\,nm on Si, 8.5\,pH/sq &
  Process spec & \url{https://seeqc.com/foundry-services} \\
\bottomrule
\end{tabular}
\end{table}

\textit{Report generated from NVB node, 2026-03-19. All quantitative claims include source URLs per Law~16.
Prices gathered 2026-03-19 --- verify before purchasing.}
